Bird beak's elaboration and exploration

### 2021-06-07 Analysis/02-MCMCglmm_mini_chains.Rmd
Update 02-MCMCglmm_mini_chains.Rmd
analyses up
Update 02-MCMCglmm_mini_chains.Rmd
Small edits for clarity and typos

All the custom code and custom functions called in this document can be downloaded and installed in `R` as the standalone working package `beer` (Beak Elaboration and Exploration in R).
Although `beer` is intentionally designed to be portable and shareable, we advice workers to use it with moderation and tailor it's consumption to their specific research needs.
Some analyses in `beer` do take some time to run so we advice to not drive or operate heavy machinery while using it.
Don't use it while driving.
The following code snippets are used to illustrate the implementation of the functions.
See §§§ for running the actual analyses.

@@

```{r load_beer, eval = TRUE}
set.seed(42)
## Loading beer
library(beer)
library(MCMCglmm)

@@

To estimate the variance covariance matrix for the phylogeny and each clade (see [01-Data_preparation.Rmd](01-Data_preparation.Rmd)) we run multivariate generalised linear mixed model (MCMCglmm) using the `MCMCglmm` package [@MCMCglmm].

@@

We ran a general multi-response model with `§§§N_traits` trait of the trait space being the the fixed effects (using a gaussian trait distribution) and and the residual covariance structure.
We used `§§§N_clades + 1` random effects (one for the whole phylogeny and one for each clade).

@@

In order to improve the speed of the analysis and take phylogenetic uncertainty into account, we used a "mini chains" approach.
This approach is not too dissimilar from @mulTree as it runs multiple MCMCglmm analyses on multiple trees and pull the results together into one larger MCMCglmm that contains more variation due to phylogenetic uncertainty.
However, the "mini chains" approach differ by the size of the chains that is optimised for speed.

@@

We first ran three independent MCMC chains with the model and data described above using the consensus tree and flat priors.
We ran these models (for `§§§N_generations` with sampling every `§§§N_nitt` and no burnin) until the three chains reached an effective sample size >> 200 (§§§and other diagnosis).

@@

We set up a mini-chain to be a MCMCglmm running with 1) the model described above, 2) a random tree and 3) the estimated priors (with a degree of parameter belief of 5% `nu = 0.05`) to run for `nitt` generations were `nitt` is equal to the previously estimated burnin phase + 100 sampled iterations (`nitt = burnin + thin * 100`).
Each mini-chain thus provided us with a 100 exploitable posterior data points.
The two main advantages of this mini-chains is that 1) they are much faster to run since no diagnoses is necessary and the chains are only ran for a relatively short time (which allow several chains to crash/fail without loosing all the data); and 2) they take into account tree uncertainty whitout having to run the complete MCMCglmm on all trees (c.f. @mulTree).

```{r, set_the_mini_chains, eval = TRUE}
## The thining parameter
thin <- 100
## The number of samples per chains
samples <- 100

@@

We then run the MCMCglmm over $N_trees$ to get an estimate of the variance co-variance matrices including phylogenetic uncertainty.

```{r, running_replicates, eval = TRUE}
## The number of replicates
n_reps <- 10

@@

#### Number of runs optimisation (TO BE DEVELOPPED?)

@@

To make sure we reached consistent results, we run the a set of N mini-chains were N was determined using the following variance optimisation algorithm

 * 1. run an initial $N = 20$ chains;
 * 2. estimate the $V_{1}$, the pooled variances for each parameters in the chain (V1 is a vector of length §§§);
 * 3. update N to N = N * 1.05 and run an additional 5% chains
 * 4. estimate $V_{2}$, the pooled variances for each parameters in the chain;
 * 5. calculate $\Delta_V$ the absolute change between $V_{1}$ and $V_{2}$ as ($\sqrt{(\frac{V_{1}}/{V_{2}} - 1)^{2}}$).
 * 6. If $\Delta_V < 0.05$ (meaning an increase of 5% of the replicates resulted in an increase of the parameters of less than 5%), the desired number of mini-chains is reached. Else, set $V_{1} = V_{2}$ and repeat from step 3.

@@

After reaching the desired $N$ chains, we pulled all the posterior estimates together back into a "normal" big MCMCglmm with a posterior distribution including the effective results and the phylogenetic uncertainty.

>>>>

All the custom code and custom functions called in this document can be downloaded and installed in `R` as the standalone working package `beer` (Beak Elaboration and Exploration in R).
Although `beer` is intentionally designed to be portable and shareable, we advise workers to use it with moderation and tailor its consumption to their specific research needs.
Some analyses in `beer` do take some time to run so we advise you do not operate heavy machinery while using it.
Don't use it while driving.
The following code snippets are used to illustrate the implementation of the functions.
See <!-- §§§run_replicates_script --> for running the actual analyses.

@@

```{r load_beer, eval = TRUE, message = FALSE}
set.seed(42)
## Loading beer
library(beer)
library(MCMCglmm)

@@

To estimate the variance covariance matrix for the phylogeny and each clade (see [01-Data_preparation.Rmd](01-Data_preparation.Rmd)) we run multivariate generalised linear mixed models (MCMCglmm) using the `MCMCglmm` package [@MCMCglmm].

@@

We ran a general multi-response model with `§§§N_traits` of the trait space as fixed effects (using a Gaussian distribution) and and the residual covariance structure.
We used `§§§N_clades + 1` random effects (one for the whole phylogeny and one for each clade).
% NC: I know this is in the data prep, but can you specify an example MCMCglmm model here to make it clearer what you mean?

@@

To increase the speed of the analysis and take phylogenetic uncertainty into account, we used a "mini chains" approach.
It runs multiple MCMCglmm analyses on multiple trees and pulls the results together into one larger MCMCglmm that contains more variation due to phylogenetic uncertainty (similar to @mulTree - ADD LINK?).
The size of the chains are optimised for speed. % NC: Editted as multree is not familiar to everyone, so keeping it general and linking to multree is better

@@

We first ran three independent MCMC chains with the model and data described above using the consensus tree and flat priors.
We ran these models (for `§§§N_generations` sampling every `§§§N_nitt` with no burnin) until the three chains reached an effective sample size >> 200 (§§§and other diagnosis). 
% NC: I thought in MCMCglmm the nitt parameter was total number of iterations, and thin gives how often the chain is sampled? I think you've called nitt generations here but better to be consistent with MCMCglmm?

@@

We set up a mini-chain to be a MCMCglmm running with 1) the model described above, 2) a random tree, and 3) the estimated priors (with a degree of parameter belief of 5% `nu = 0.05`) to run for `nitt` generations were `nitt` is equal to the previously estimated burnin phase + 100 sampled iterations (`nitt = burnin + thin * 100`).
Each mini-chain thus provided us with 100 posterior data points.
The two main advantages of this mini-chain approach is that 1) they are much faster to run since no diagnosis of convergence is necessary and the chains are only run for a relatively short time (which allow several chains to crash/fail without losing all the outputs); and 2) they take into account tree uncertainty without having to run the complete MCMCglmm on all trees (c.f. @mulTree).

```{r, set_the_mini_chains, eval = TRUE}
## The thinning parameter
thin <- 100
## The number of samples per chains
samples <- 100

@@

We then run the MCMCglmm over $N_trees$ to get an estimate of the variance covariance matrices including phylogenetic uncertainty.

```{r, running_replicates, eval = FALSE}
## The number of replicates
n_reps <- 10

@@

<!--

#### Number of runs optimisation (TO BE DEVELOPED?)

@@

To make sure we reached consistent results, we run the set of N mini-chains were N was determined using the following variance optimisation algorithm.

 * 1. run an initial $N = 20$ chains;
 * 2. estimate the $V_{1}$, the pooled variances for each parameters in the chain (V1 is a vector of length §§§);
 * 3. update N to N = N * 1.05 and run an additional 5% chains;
 * 4. estimate $V_{2}$, the pooled variances for each parameters in the chain;
 * 5. calculate $\Delta_V$, the absolute change between $V_{1}$ and $V_{2}$ as ($\sqrt{(\frac{V_{1}}/{V_{2}} - 1)^{2}}$).
 * 6. If $\Delta_V < 0.05$ (meaning an increase of 5% of the replicates resulted in an increase of the parameters of less than 5%), the desired number of mini-chains is reached. Else, set $V_{1} = V_{2}$ and repeat from step 3.

@@

After reaching the desired $N$ chains, we pulled all the posterior estimates together back into a "standard" big MCMCglmm with a posterior distribution including the effective results and the phylogenetic uncertainty.

-->
